R Notebook

### 2017-03-21 manuscript/draft.Rmd

* BBS + retriever
* Predictor variables
  * Monthly climate data from interpolated weather stations (PRISM), converted
    to BIOCLIM variables. Many of these variables were tightly correlated, so
    we only included the 8 variables selected by Harris's (2015) BBS analysis
  * NDVI (GIMMS)
  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
* Data summary: 
  * Number of sites, species, unique observers
  * central moments of richness (mean about 51, sd about 12 spp)
  * The within-site variation was much smaller (sd about 5 spp), making the
    site-level distributions underdispersed.  
    * This underdispersion implies
      that common count models such as binomial, Poisson, negative binomial 
      won't fit well.
    * For this reason (and because the time series methods assumed 
      Gaussian errors), we defined model performance throughout this paper in 
      terms of Gaussians (e.g. mean squared error), rather than with count
      distributions.
        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
          distributions internally, however.

@@

**Time series models**
* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
  model selection based on AIC.
* Auto-ARIMA + environment.
* Uncertainty treated as above

**"Macroecological" model**
* Use a nonlinear model to predict richness as a function of environmental predictors
  (BIOCLIM + elevation + NDVI).
* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
  noisy species-level estimates, but doesn't use the full data set in one model.
* For this part of the analysis, we used the `gbm` package for fitting boosted regression
  trees with Gaussian errors.  
* We included observer effects & the associated uncertainty by running this model [[N]]
  times (once per MCMC sample).
* We did *not* include the site-level random effects as predictors, meaning that this
  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
  NDVI values should have identical richness distributions.

**SDMs**
* Probably the most common approach?
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we
  used a custom wrapper around the randomForest package. Random forests are flexible
  models that are built by combining many independent models ("trees") fit to different 
  versions of the training data, which made it straightforward to adapt them to our 
  situation: we simply fit one tree using each of the Monte Carlo estimates of our 
  observer effects and averaged the predictions of the resulting forest.

**JSDM** 
* New approach that tries to get the , cite examples.
* Should get best of the previous 2. Might be able to propagate uncertainty
  from lme4.

@@

Evaluated each year after the last training year (2010?):
* (R)MSE 
  * How far, on average, are the models' predictions from the true value?
  * [[RMSE has better units than MSE (species versus squared species), but MSE is
    nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
* Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
  and coverage)
* Coverage at 95%
* Calibration/bias

>>>>

* BBS + retriever
* Predictor variables
  * Monthly climate data from interpolated weather stations (PRISM), converted
    to BIOCLIM variables. Many of these variables were tightly correlated, so
    we only included the 8 variables selected by Harris's (2015) BBS analysis
  * Weather/climate and NDVI were each calculated separately for each year,
    rather than averaged over the full multi-decade period (as Worldclim does)
  * NDVI (GIMMS)
  * Altitude (SRTM 90m Digital Elevation Database via raster::getData)
* Data summary: 
  * Number of sites, species, unique observers
  * central moments of richness (mean about 51, sd about 12 spp)
  * The within-site variation was much smaller (sd about 5 spp), making the
    site-level distributions underdispersed.  
    * This underdispersion implies
      that common count models such as binomial, Poisson, negative binomial 
      won't fit well.
    * For this reason (and because the time series methods assumed 
      Gaussian errors), we defined model performance throughout this paper in 
      terms of Gaussians (e.g. mean squared error), rather than with count
      distributions.
        * Some of the modeling approaches (SDMs and JSDMs) did use discrete
          distributions internally, however.
* Splitting the data temporally
  * Our data set began in `r settings$start_yr`, which was the first year that 
    NDVI data was available from GIMMS
  * All the models were trained on the data that was collected through 
    `r settings$last_train_year`
  * Sites were included in the analysis if there were at least 
    `r settings$min_num_yrs` years of data available during this period.
  * The models were tested/evaluated on the data from 
    `r settings$last_train_year + 1` `r settings$end_yr`.

@@

**Time series models**
* These models only looked at one site at a time and looked for patterns in 
  univariate time series.
* Auto-ARIMA (Hyndman 2016). Choose a differencing order, then do forward stepwise
  model selection based on AICc.
    * Autoregressive: current state depends on previous state(s)
    * Moving average: current observation error depends on previous observation error
    * Integrated/differencing: is analysis performed on raw data, on sequential 
      differences of the raw data, on sequential differences of sequential differences,
      etc.
    * default settings, except seasonal = FALSE
* Auto-ARIMA + site-level environmental variables over time.
* Uncertainty treated as above

**"Macroecological" model**
* Use a nonlinear model to predict richness as a function of environmental predictors
  (BIOCLIM + elevation + NDVI).
* We chose the `gbm` package for fitting boosted regression trees as our example of
  a nonlinear "macroecological model".
    * Works by fitting thousands of small models sequentially, and then adding up
      their predictions
    * the individual small models were regression trees with [[settings]].
* Compared to SDMs, this lets us predict richness directly instead of using hundreds of
  noisy species-level estimates, but doesn't use the full data set in one model.
* For this part of the analysis, we used the `gbm` package for fitting boosted regression
  trees with Gaussian errors.  
* We included observer effects & the associated uncertainty by running this model [[N]]
  times (once per MCMC sample).
* We did *not* include the site-level random effects as predictors, meaning that this
  approach implicitly assumes that two sites with identical Bioclim, elevation, and 
  NDVI values should have identical richness distributions.

**SDMs**
* Probably the most common approach?
* These models say that species' occurrence probabilities depend only on the values of
  the environmental variables at the time and place of the transect, not on the history
  of the environment or the history of the birds at that site
* Predict each species' occurrence probabilities individually, then add up the predictions
* [[Something about not using thresholds that throw away information]]
* Taking this approach in the face of uncertainty about observer effects would either 
  require a customized model (cite "neighborly advice" paper), or fitting [[N]] models
  times the number of species.
* To make it feasible to fit this large number of models, we used a custom wrapper 
  around the randomForest package. Random forests are flexible models that are built 
  by combining many independent models ("trees") fit to different versions of the training
  data, which made it straightforward to adapt them to our situation: we simply fit one 
  tree using each of the Monte Carlo estimates of our observer effects and averaged the 
  predictions of the resulting forest.
* For the uncertainty, we used Gaussian approximations, based on the means and variances
  of the sums of independent Bernoulli random variables (cite Calabrese et al.).

**JSDM** 
* Joint species distribution model (JSDM) is a new approach that makes predictions
  about the full community composition, rather than modeling it species-by-species.
* In stacked single-species SDMs, the species-level occurrences are treated as 
  independent coin-flips whose proability depends only on the environment.
* JSDMs remove this independence assumption, and explicitly account for the
  possibility that a site will be much more (or less) suitable for birds in 
  general than one would expect based on a few environmental measurements alone.
* As a result, JSDMs do a better job of representing our uncertainty about 
  richness, while stacked SDMs underestimate it (Harris 2015).
* We used the `mistnet` package (Harris 2015) because it is the only JSDM that
  describes species' environmental associations with nonlinear functions.
* The mistnet package doesn't "know" about time series or spatial autocorrelation,
  so it did not share information among the repeated runs of the same transect.

@@

* Evaluated each year between  `r settings$last_train_year` and `r settings$end_yr`
* Metrics:
  * (R)MSE 
    * How far, on average, are the models' predictions from the true value?
    * [[RMSE has better units than MSE (species versus squared species), but MSE is
      nice because it's additive (i.e. MSE of 2 is twice as big as MSE of 1).]]
  * Calibration/bias
    * In general, or in specific situations, is the model consistently predicting
      that richness will be too high or too low?
    * Can also report this value in terms of RMSE or MSE
  * Coverage of 95% confidence intervals
  * Gaussian Log-likelihood (simultaneously rewards good point estimates, precision, 
    and coverage)
